Covariate analysis of ROSMAP RNASeq

### 2016-01-29 code/Rmd/Mayo_Cerebellum_Array.Rmd

# Query biomart
Illumina2HGNC <- getBM(attributes = c("illumina_humanht_12_v4", "hgnc_symbol"),
                       filters = "illumina_humanht_12_v4", 
                       values = RESIDUAL.GENE_EXPRESSION$illumina_humanht_12_v4,
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(Illumina2HGNC, RESIDUAL.GENE_EXPRESSION) %>%
  dplyr::select(illumina_humanht_12_v4, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB Cerebellum Microarray', parentId = parentId)
CODE <- synStore(CODE)

@@

# Store expression values
write.table(RESIDUAL.GENE_EXPRESSION, file = 'expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('expression.tsv', name = 'Voom Adjusted Residual Expression (Dxn added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
```
|  *Results*            |  *SynapseID*                    |
|  -------              |   ---------                     |
|  Covariates           |  `r COV_OBJ$properties$id`      |
|  Residual Expression  |  `r EXP_OBJ$properties$id`      |

>>>>

# Query biomart
Illumina2HGNC <- getBM(attributes = c("illumina_humanht_12_v4", "hgnc_symbol"),
                       filters = "illumina_humanht_12_v4", 
                       values = RESIDUAL.GENE_EXPRESSION$illumina_humanht_12_v4,
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(Illumina2HGNC, RESIDUAL.GENE_EXPRESSION) %>%
  dplyr::select(illumina_humanht_12_v4, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp}
# Get design matrix
DM = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F], Intercept = F)

# Linear model with new design
FIT = lmFit(EXPR, design=DM$design)

# Fit contrast
contrast = makeContrasts(contrasts="Dxn0-Dxn1", levels = colnames(FIT$coefficients))
FIT = contrasts.fit(FIT, contrasts=contrast)
FIT = eBayes(FIT)

# Get differnetial expression
DE = topTable(FIT, coef=1, number = dim(EXPR)[1])
DE = rownameToFirstColumn(DE, 'illumina_humanht_12_v4') %>%
  left_join(Illumina2HGNC)
p <- ggplot(DE, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
p

tmp = DE %>% dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(hgnc_symbol) %>%
  unlist %>% unique
kable(data.frame(Comparison = "Control_vs_AD",
           nDiffExp = length(tmp),
           DiffExpGenes = paste(tmp, collapse = ',')))  
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB Cerebellum Microarray', parentId = parentId)
CODE <- synStore(CODE)

@@

# Store expression values
write.table(RESIDUAL.GENE_EXPRESSION, file = 'expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('expression.tsv', name = 'Voom Adjusted Residual Expression (Dxn added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'diffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('diffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

```
|  *Results*            |  *SynapseID*                    |
|  -------              |   ---------                     |
|  Covariates           |  `r COV_OBJ$properties$id`      |
|  Residual Expression  |  `r EXP_OBJ$properties$id`      |
|  Differential Expression  |  `r DEXP_OBJ$properties$id`      |

### 2016-01-29 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA <- METADATA %>%
  dplyr::filter(!(ID %in% c("1923_TCX")))
```
Following five samples were considered outliers and removed 1923\_TCX
```{r preprocess.data1}
indToRetain = intersect(METADATA$ID, colnames(COUNT))

@@

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)
```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=NULL, plot=T)

# Initial design matrix with covariates under consideration
DM = getDesignMatrix(COVARIATES, Intercept = F)
```
Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data, fig.height=9, fig.width=16, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p

# Tree based clustering
sampleTrees = flashClust::hclust(dist(t(VOOM.GENE_EXPRESSION$E)), method = "average")
tree = WGCNA::cutreeStatic(sampleTrees, cutHeight = 200, minSize=3)

@@

# Recalculate weights with new design
VOOM.ADJUSTED.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, 
                                     design=DM1$design, plot=T)

@@

### Clustering residual gene expression
```{r decompse.normalise.data1, fig.height=10, fig.width=16, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1)], scale.=T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

# Tree based clustering
sampleTrees = flashClust::hclust(dist(t(RESIDUAL.GENE_EXPRESSION[,-(1)])), method = "average")

@@

>>>>

### Data preprocessing
```{r preprocess.data,cache=FALSE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA <- METADATA %>%
  dplyr::filter(!(ID %in% c("1923_TCX")))
```
Following samples were considered outliers and removed based on PCA and/or eucledian distance: 1923\_TCX
```{r preprocess.data1}
indToRetain = intersect(METADATA$ID, colnames(COUNT))

@@

### Normalisation (with NULL)
Initial normalisation usign voom (with NULL design)

```{r initial.voom.normalisation}
# Initial normalisation of gene expression
VOOM.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, design=NULL, plot=T)

# Initial design matrix with covariates under consideration
DM = getDesignMatrix(COVARIATES, Intercept = F)
```

Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data1, fig.height=10, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p
```
```{r decompse.normalise.data2, fig.height=9, fig.width=25, results='asis'}
# Tree based clustering
sampleTrees = flashClust::hclust(dist(t(VOOM.GENE_EXPRESSION$E)), method = "average")
tree = WGCNA::cutreeStatic(sampleTrees, cutHeight = 200, minSize=3)

@@

# Recalculate weights with new design
VOOM.ADJUSTED.GENE_EXPRESSION = voom(EXPR$filteredExprMatrix, 
                                     design=DM1$design, plot=F)

@@

### Clustering residual gene expression
```{r decompse.normalise.data3, fig.height=10, fig.width=9, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1)], scale.=T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
```{r decompse.normalise.data4, fig.height=10, fig.width=25, results='asis'}
# Tree based clustering
sampleTrees = flashClust::hclust(dist(t(RESIDUAL.GENE_EXPRESSION[,-(1)])), method = "average")

@@

### Source R Code
[This File](`r thisFile`)

### 2016-01-29 code/Rmd/ROSMAP_RNASeq.Rmd

## Load required libraries
library(synapseClient)
library(ggplot2)
library(reshape2)
library(data.table)
library(dplyr)
library(stringr)
library(limma)
library(knitr)
library(Biobase)
library(GEOquery)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(ggplot2)

@@

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION[,-(1:2)])))
plotDendroAndColors(tree, 
                    colors = labels2colors(COVARIATES.tmp), 
                    dendroLabels = FALSE, 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES.tmp))
```

@@

DE = DE %>% plyr::ldply(.id = 'Comparison')

>>>>

## Load required libraries
library(synapseClient)
library(ggplot2)
library(reshape2)
library(data.table)
library(dplyr)
library(stringr)
library(limma)
library(knitr)
library(Biobase)
library(GEOquery)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(ggplot2)
library(WGCNA)

@@

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION[,-(1:2)])))
WGCNA::plotDendroAndColors(tree, 
                    colors = labels2colors(COVARIATES.tmp), 
                    dendroLabels = FALSE, 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES.tmp))
```

@@

DE = DE %>% plyr::ldply(.id = 'Comparison') %>%
  tidyr::separate(ensembl_gene_id, into=c("ensemble_id","position")) %>%
  dplyr::rename(ensembl_gene_id = ensemble_id) %>%
  left_join(Ensemble2HGNC)
